---
title: "Code-along, Week 10: Modelling pubs, continued, continued"
author: "Alex Homer"
date: "25 November 2021"
output: github_document
editor_options: 
  chunk_output_type: console
---

This week's data come from the UK's [Office for National Statistics](https://www.ons.gov.uk/) (ONS).  We will be using the following datasets:

* [Public houses and bars by local authority](https://www.ons.gov.uk/businessindustryandtrade/business/activitysizeandlocation/datasets/publichousesandbarsbylocalauthority) (Copyright date: 2018)
* [Estimates of the population for the UK, England and Wales, Scotland and Northern Ireland ](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/populationestimatesforukenglandandwalesscotlandandnorthernireland) (Copyright date: 2021)
* [Earnings and hours worked, place of residence by local authority: ASHE Table 8](https://www.ons.gov.uk/employmentandlabourmarket/peopleinwork/earningsandworkinghours/datasets/placeofresidencebylocalauthorityashetable8) (Copyright date: 2021)
* [Standard Area Measurements (2018) for Administrative Areas in the United Kingdom](https://geoportal.statistics.gov.uk/datasets/standard-area-measurements-2018-for-administrative-areas-in-the-united-kingdom/about) (Copyright date: 2019)
* [Life Expectancy by Local Authority](https://www.ons.gov.uk/datasets/life-expectancy-by-local-authority/editions/time-series/versions/1) (Copyright date: 2020)
* [Local Authority Districts (December 2018) Boundaries UK BFC](https://geoportal.statistics.gov.uk/maps/fef73aeaf13c417dadf2fc99abcf8eef/about) (Copyright date: 2019)

The data are subject to crown copyright and database rights and are used under the [Open Government Licence v3.0](https://www.nationalarchives.gov.uk/doc/open-government-licence/version/3/).  In addition the boundary file contains OS data © Crown copyright and database right 2019.

```{r load-packages, message = FALSE}
library(tidyverse)
library(tidymodels)
library(sf)
library(here)
```

```{r setup, include = FALSE}
lightbulb <- function(x) {
  paste(emo::ji("bulb"), " *", x, "* ", emo::ji("bulb"), sep = "")
}
```

## Read data

I've already done some of the cleanup and joining in advance to save time in the live session.  You can see the cleanup script in the "data" folder of this repo.

```{r read-data, message = FALSE}
pubs_data <- readRDS(here("data/pubs-final.rds"))
```

Because I put these data together, let's write a data dictionary:

- `area_code`: the [ONS area code](https://en.wikipedia.org/wiki/ONS_coding_system) for each district
- `area_name`: the name of each district
- `num_pubs`: the number of pubs in each district (2018)
- `pop`: the population of each district (2018)
- `pubs_per_capita`: the number of pubs per person (2018; obtained by dividing `num_pubs` by `pop`)
- `country`: which UK nation each district belongs to (England, Northern Ireland, Scotland or Wales)
- `median_pay_2017`: the median weekly pay, in pounds sterling, for residents of each district (2017)
- `area_sqkm`: the area, in square kilometres, of each district to mean high water level
- `coastal`: categorical variable, with three values:
  - `"Inland"`, for districts in Great Britain whose total extent is equal to their extent to mean high water level
  - `"Coastal"`, for districts in Great Britain whose total extent is greater than their extent to mean high water level
  - `NA`, for districts in Northern Ireland (the data in NI appear to have been measured differently, so all districts have equal values for their total extent and extent to mean high water)
- `pop_dens`: the population density (people per square kilometre) of each district (2018; obtained by dividing `pop` by `area_sqkm`)
- `life_exp_female`, `life_exp_male`: the life expectancy at birth of people born in each district, broken down by binary sex (2016--2018) ^[Sex and gender are, of course, more complicated than that, but the ONS only gives "male" and "female" figures; in the UK all birth certificates have one of these two options recorded.]

It's worth noting that the `coastal` variable doesn't necessarily agree with intuitive definitions of the words "coastal" and "inland", because of the effect of tidal rivers on the data...

```{r map, cache = TRUE, fig.height = 10}
shape_data <- read_sf(
  here(
    "data/map/Local_Authority_Districts_(December_2018)_Boundaries_UK_BFC.shp"
  )
)

shape_data %>%
  rename(area_code = lad18cd) %>%
  left_join(pubs_data, by = "area_code") %>%
  # Filter for Scotland so it knits on RStudio Cloud; remove this next line if
  # you want to make the whole map and you can get the code to run!
  filter(str_detect(area_code, "^S")) %>%
  ggplot(aes(fill = coastal)) +
  geom_sf(colour = "black") +
  scale_fill_manual(values = c(Coastal = "turquoise", Inland = "darkGreen")) +
  labs(fill = "Area type") +
  theme_minimal()
```

By this definition, Glasgow is coastal, but Glaswegians might disagree with that!


## Exploratory data analysis

### Missing data

Let's have a look at the variables we have.

```{r glimpse}
glimpse(pubs_data)
```

It would be good to see where our missing data are.  For this, there are two functions you haven't necessarily seen before: `across` and `select_if`.  `across` lets you apply a function to multiple columns, and `select_if` lets you select columns that satisfy a certain condition.

In both cases we use R's tilde (`~`) and dot (`.`) notation to specify the function we want to apply (we can think of a logical condition as a special kind of function that outputs the values `TRUE` or `FALSE`).

```{r see-NAs}
cols_with_NAs <- pubs_data %>%
  summarise(across(everything(), ~sum(is.na(.)), .names = "NAs_{.col}")) %>%
  select_if(~ sum(.) > 0)

cols_with_NAs
```

We'd like to see which areas have `NA`s against them.  Since we'll be applying the same operations repeatedly, we can write a function for this, whose argument will be a column name.

```{r this}
which_nas <- function(var_name) {
  pubs_data %>%
    filter(is.na((!!as.symbol(var_name)))) %>%
    pull(area_name)
}

col_choices <- cols_with_NAs %>%
  colnames() %>%
  str_remove("NAs_")

NA_places <- map(col_choices, which_nas) %>%
  set_names(col_choices)

NA_places
```

What we get here is a named list, where each element of the list is a vector of places that have the value `NA` in one of the columns, and the name of each list element is the name of a column some of whose elements are `NA`.

We can see here that:

- The median pay is unavailable for the City of London and the Isles of Scilly---unsurprising, as these are very small areas in terms of population---and for the eleven Northern Irish local authorities, presumably because this element of data collection is devolved to the Northern Ireland Executive.
- The eleven NI areas are also those where the coastal/inland data is unavailable, as referenced in the data dictionary.
- The life expectancy data is unavailable for the City of London and the Isles of Scilly, again, and also for various English local authority areas that have been abolished since 2018.  This is because the data, while applying to 2016--18, have been calculated for the current council areas and not for the ones that existed at the time.

### Pairs plot

A quick way to explore multiple variables at once is to do a pairs plot, which is provided by the `ggpairs` function from the `GGally` package.

```{r ggpairs, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 10, cache = TRUE}
pubs_data %>%
  select(-c(area_code, area_name, coastal, num_pubs, pop)) %>%
  GGally::ggpairs(aes(colour = country))
```

It's a bit of a messy plot, and I've had to use both `message = FALSE` and `warning = FALSE` in the chunk options above to avoid printing a load of junk to the output.  I've advised in the past that putting `message = FALSE` is generally fine, but usually when you're tempted to use `warning = FALSE` you should instead fix whatever is causing the warning to be displayed.  So why am I going against my own advice?

The reason is that this is a plot that's more useful for your own explorations than for you to present to others; you would probably want to select some of the plots to present, since it's hrd to tell a coherent story with this on its own.  Meanwhile it would take quite a while to fix a lot of the issues, and that probably isn't worth it for a plot just for yourself.

We're particularly interested in the first column, showing relationships between `pubs_per_capita`, our intended response variable, and the other variables.  But it's hard to see because there are two outliers that squash most of the points into the extreme left of the plot; they don't show up everywhere because of missing data.  We can work out which these are and remove them.

`r lightbulb("Before reading the output of the code below, can you work out which two places they must be, given that they don't appear in the plots for median pay and for life expectancy?")`

```{r outliers}
pubs_data %>%
  slice_max(pubs_per_capita, n = 2) %>%
  select(area_name, num_pubs, pop, pubs_per_capita)
```

It's our old friends, the City of London and the Isles of Scilly.  These areas have particularly low populations, which force the number of pubs per capita upwards.  In addition, the City of London has a large non-resident population---over half a million people work there ^[Source: [City of London website](https://www.cityoflondon.gov.uk/supporting-businesses/economic-research/statistics-about-the-city)], compared to the nine thousand or so who live there---who presumably frequent its pubs.  Meanwhile, the Isles of Scilly consist of a number of [separate islands without bridges between them](https://www.visitislesofscilly.com/explore/island-hopping), that are also separated from the mainland.  So I think it's fair to say that these two observations are sufficiently distinct from the others that we can remove them as outliers.

`r lightbulb("Why did I need to justify removing the outliers?  What stopped me from just removing them?")`

Let's create a slightly-reduced dataset without these two points, and repeat the `ggpairs` plot.

```{r pubs-reduced}
pubs_reduced <- pubs_data %>%
  filter(!(area_name %in% c("City of London", "Isles of Scilly")))
```

`r lightbulb("I could have done this by filtering for a pubs-per-capita figure that was higher than 0.004.  Why didn't I?")`

```{r ggpairs-fixed, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 10, cache = TRUE}
pubs_reduced %>%
  select(-c(area_code, area_name, coastal, num_pubs, pop)) %>%
  GGally::ggpairs(aes(colour = country))
```

That's better!

`r lightbulb("One Scottish observation seems to have a very high area: that's the Highland council area in Scotland.  Can we remove that as an outlier too?")`

## Modelling

### Univariate regression

It looks like there's a negative vaguely-linear trend on the graph of pubs per capita vs pay.  Let's model that.

```{r pay-model}
pubs_reduced <- pubs_data %>%
  filter(!(area_name %in% c("City of London", "Isles of Scilly")))

mod <- linear_reg() %>%
  set_engine("lm") %>%
  fit(pubs_per_capita ~ median_pay_2017, data = pubs_reduced)

tidy(mod)
```

```{r extract-fun, include = FALSE}
abs_str <- function(mod_input, term_input, scale = 1) {
  mod_input %>%
    tidy() %>%
    filter(term == term_input) %>%
    pull(estimate) %>%
    magrittr::multiply_by(scale) %>%
    signif(digits = 3) %>%
    format(scientific = FALSE)
}
```

We can interpret this as saying that:

- for a district with a median pay of £0 per week, the predicted number of pubs per capita is `r abs_str(mod, "(Intercept)")`, or one pub per thousand people.
- for every pound per week that median pay increases, the predicted number of pubs per capita *decreases* by `r abs_str(mod, "median_pay_2017")`.

That last number isn't very human readable.  Let's consider instead the number of pubs per 10,000 people.  Let's also consider what happens when pay increases by £100 per week.  This has the combined effect of multiplying the number by $10\,000 \times 100 = 1\,000\,000$.  So we have that

- for every £100 per week that median pay increases, the predicted number of pubs per 10,000 people *decreases* by `r abs_str(mod, "median_pay_2017", scale = 1e6)`.

`r lightbulb("Is the intercept meaningful here?  Note that, since pay can't be negative, a district with median pay of £0 per week must have more than half of its population earning £0 per week.")`

```{r pay-model-2}
glance(mod)
```

The $R^2$ isn't amazing here.  Perhaps we can improve the model fit by considering more terms?

### Adding a categorical predictor

This document will be completed during the live session.

```{r workflow}
pubs_rec <- recipe(pubs_per_capita ~ ., data = pubs_reduced) %>%
  step_rm(area_name, area_code, num_pubs, pop, area_sqkm) %>%
  step_filter(country != "Northern Ireland") %>%
  step_dummy(all_nominal())

pubs_mod <- linear_reg() %>%
  set_engine("lm")

pubs_wflow <- workflow() %>%
  add_model(pubs_mod) %>%
  add_recipe(pubs_rec)
```

```{r test-train-split}
set.seed(8651)
pubs_split <- initial_split(pubs_reduced, prop = 0.8, strata = country)
pubs_train <- training(pubs_split)
pubs_test  <- testing(pubs_split)

pubs_train_fit <- pubs_wflow %>%
  fit(data = pubs_train)
tidy(pubs_train_fit)
glance(pubs_train_fit)
```